---
title: "Schoolwork and Personal Time Across Weekdays and Weekends at Amherst"
subtitle: "STAT 231: Calendar Query"
author: "Sebastian Montesinos"
date: "Last updated \\today"
output: pdf_document
fig_width: 6
fig_height: 4
---

```{r setup, include = FALSE}
# load packages
library(tidyverse)
library(lubridate)
library(ical)
library(kableExtra)
# set code chunk defaults
knitr::opts_chunk$set(tidy = F, # display code as typed
                      size = "small", # slightly smaller code font
                      message = FALSE,
                      warning = FALSE,
                      comment = "\t") 

# set black & white default plot theme 
# theme_set(theme_classic()) # uncomment this line if desired

# improve digit and NA display 
options(scipen = 1, knitr.kable.NA = '')
```

<!--
HOW TO USE THIS TEMPLATE:

The template is designed to help you structure your report. All work herein should be your own. Thus, you should write your own code comments, set your own chunk options, etc. You should clean up code comments that I've left so they are useful for you. 

You can leave the markdown comments for guidance in writing your report since they will not appear in the knitted pdf. If you like, you can remove them. 

Please rename/remove the headings as you see fit. 
You may add subheadings as desired to break up the report. Practice your Markdown formatting to make a professional looking report. 

REMINDERS ABOUT FILE ORGANIZATION AND GITHUB:
You should commit and push all relevant files to GitHub. Someone else (me, in particular) should be able to download the *calendar-query* folder from GitHub and run your .Rmd file without changing anything (meaning file paths are not specific to your computer and all files that are read in are provided.)

Your final *calendar-query* project folder should be organized such that file names and any subfolder names are informative, making relevant content easy to find for someone with little or no context of the project (i.e., if someone saw a file called "calendar_report_template" they would likely assume it was a template and *not* and not a finalized report). 
-->

# Introduction 

<!--
Describe the questions of interest: what are they and why are they important or useful to you?
-->

The first question I addressed is how the amount of time I spend on each class changes over the course of a week. One of my goals this semester is to have a structured and planned week where I dedicate certain days to certain classes. This allows me to easily plan when I will do work and not get overloaded on any particular day. I aim to concentrate my data science work towards the end of the week - specifically, around Thursday, Friday, and Saturday. I aim to do my oceanography and religion work around the start of the week, particularly Monday and Tuesday respectively. Finally, I aim to scatter my thesis work throughout the week. 

The second question I address is how the amount of time I spend on work in a day relates to the amount of personal time I take during a day. One might expect that since the more time I spend on one activity the less time there is for the other, there will be a simple negative correlation. However, I have noticed that I sometimes give myself more time for recreational activities when I have done a lot of work as a 'reward', and spend more time on independent academic projects or looking for jobs when I do not spend a lot of time on work in a day since I feel guilty if I am not productive in a day. Therefore, it could also be the case that as I spent more time on schoolwork I also take more personal time, making the two positively correlated. I was also interested in how this relationship might change on weekdays vs weekends, since I try to give myself more of a break during the weekends.

The final question I addressed was how much time I spend doing work on weekdays versus on weekends. One of my aims at Amherst has been to have concentrated, productive weekdays so that I mostly have my weekends off. However, since I am writing a thesis this semester that has been harder to accomplish. I was interested in how productive I managed to be during the weekdays, and whether the time I spent on work during the weeks differed from on the weekends.

# Data collection 

<!--
Briefly describe your data collection process, including how you defined variables of interest (include the levels/categories of categorical variables and the units for quantitative variables).
--> 
I coded my data into three broad categories: 'work', for anything related to school (ie. class and homework), 'personal' for anything related to recreation (ie. gaming, seeing friends) and 'extracurricular' for any club activities. I decided to code the work category as four subcategories corresponding to my four classes: data science, oceanography, religion, and my thesis. I decided not to further divide the other two categories since my questions did not require doing so. So, in the end, I had six ways I marked off time on my calendar: 4 for my classes, 1 for personal time, and 1 for extracurricular. The units for all of these categories were in time, specifically minutes spent on each activity, which I used google calendar to code in. I did not code every single thing I did in the day, so tasks such as job applications or independent projects did not show up on my calendar.


<!--
The code below is provided as a starting point for importing an `ics` file into R as a dataframe, doing some initial wrangling, and adding up the time spent on each activity. You are not required to use this code, but I hope it is useful as you get started. Feel free to delete or modify it as you see fit. 

If you are using the code, be sure to update the capitalized parts with the relevant path to where your `ics` file is located *relative to where this Rmd file is located in the repo*, and set `eval = TRUE`.

NOTE: All of your wrangling should be in this file - preferably well-organized and in one place if possible. You may use multiple chunks if it makes it easier to tell your narrative. This is demo-ed below, where the data is wrangled first for one visualization, then could be used for the second, etc. 
-->

```{r import-wrangle, eval = TRUE}
# Data import and preliminary wrangling
calendar_data <- "SMontesinosCalendarQuery.ics" %>% 
  ## Use ical package to import into R
  ical_parse_df() %>% 
  ## Convert to "tibble" data frame format
  as_tibble() %>% 
  ## calendar event descriptions are in a variable called "summary"
  ## "activity" is a more relevant/informative variable name
  rename(activity = summary) %>% 
  mutate(
    ## Specify time zone (defaults to UTC otherwise)
    start_datetime = with_tz(start, tzone = "America/New_York"),
    end_datetime = with_tz(end, tzone = "America/New_York"),
    ## Compute duration of each activity in hours
    ## Feel free to use minutes instead
    duration = interval(start_datetime, end_datetime) / hours(1),
    ## Convert text to lower case and trim spaces to help clean up
    ## potential inconsistencies in formatting
    activity = str_to_lower(activity),
    ## separate date from time
    date = floor_date(start_datetime, unit = "day"),
    ## Parsing dates and times
    year = year(date),
    month = month(date, label = FALSE),
    day = day(date),
    day_of_week = wday(date, label = TRUE),
    day_of_year = yday(date)) %>% 
  ## remove spurious year (added to every Google calendar)
  filter(year != 1969) %>%
  ## Turning the date variable into a date type
  mutate(date = ymd(date)) %>%
  ## Including only dates after I started collecting for the project
  filter(year >= 2022 & month >= 2 & day >= 17 | year >= 2022 & month >= 3) %>%
  ##Removing trailing whitespace
  mutate(activity = str_trim(activity)) %>%
  ##Replacing spaces with underscores in activity names 
  mutate(activity = str_replace(activity, " ", "_")) %>%
  ##Creating another column that records whether the observation was made on the weekday or weekend
  mutate(week_status = case_when(day_of_week == "Sat" | day_of_week == "Sun" ~ "Weekend", 
                                 TRUE ~ "Weekday"))

```

<!--
Briefly describe and justify your plans for analysis (i.e., visualizations and summary table). Which graphs and corresponding visual cues do you plan to use? Which summary measures will you use in your table? (This will likely be a less detailed version of what you submitted for your proposal.) The idea is to let your reader know what is coming before they get to it. 

--> 

To address my first question, I intend to create a line graph that shows the average amount of time I spend on work on each day of the week, by the specific class I am working on. This graph will have the day of the week on the x axis and the time I spend on work on the y axis. There will be four lines, each corresponding to my four courses. 

To address my second question, I intend to create a scatterplot that shows the relationship between the amount of work I do in a day and the amount of personal time I take in a day. The x axis will represent work time and the y axis will represent personal time. A line of best fit will be used if appropriate, and I will use faceting to create a version of the graph for weekdays & for weekends.

To answer my final question, I intended to create a table that displays a variety of summary statistics about the time I spend on the weekdays vs weekends. Specifically, I intended to show the mean, median, and standard deviation of my time spent on schoolwork, personal activities, and extracurriculars on weekdays vs weekends.






```{r vis1-prep, eval = TRUE}
# Preparing dataset for first visualization
# Computing total duration for each activity per date
activities_total <- calendar_data %>% 
  group_by(date, day_of_week, activity) %>%
  summarize(duration = sum(duration)) %>%
#Pivoting wider to get rows as days and activities as columns
  pivot_wider(names_from = activity, values_from = duration)

#Filling in columns with NAs with '0' for 0 time spent
activities_total[is.na(activities_total)] = 0

#Calculating the mean time spent for each activity by day of the week
activities_average <- activities_total %>%
  group_by(day_of_week) %>%
  summarise(DS_Average = mean(data_science), 
            O_Average = mean(oceanography), 
            R_Average = mean(religion),
            T_Average = mean(thesis)) %>%
#Pivoting back longer to put the mean time for each activity in one column
  pivot_longer(-day_of_week, 
               names_to = "Activity",
               values_to = "Duration")

```

```{r vis2-prep, eval = TRUE}
# Preparing dataset for second visualization
activities_comparison <- calendar_data %>%
#Selecting relevant variables
  select(day, activity, duration, week_status) %>% 
# Adding a unique row identifier so I can pivot
  mutate(row = row_number()) %>%
# Pivoting wider to get each activity by day
  pivot_wider(names_from = activity, values_from = duration) %>%
# Dropping row identifier
  select(-row) 
#replacing NA values with 0s
activities_comparison[is.na(activities_comparison)] = 0

# Grouping by day and week status
activities_comparison2 <- activities_comparison %>%
  group_by(day, 
           week_status) %>%
# Calculating total school work time, personal time, and extracurricular time for each day
  summarise(work_time = sum(thesis, 
                            oceanography, 
                            religion, 
                            data_science), 
            personal_time = sum(personal), 
                            extracurricular_time = sum(extracurriculars))
# 




```


```{r vis3-prep, eval = TRUE}
# Preparing dataset for table

#Grouping adjusted dataset from Vis2-prep by only week_status
week_comparison <- activities_comparison2 %>%
  group_by(week_status) %>%
#Calculating the mean, median, and standard deviation of work, personal, and extracurricular time by week status
  summarise("Mean Work Time" = mean(work_time), 
            "Median Work Time" = median(work_time), 
            "Standard Deviation (Work Time)" = sd(work_time),
            "Mean Personal Time" = mean(personal_time),
            "Median Personal Time" = median(personal_time),
            "Standard Deviation (Personal Time)" = sd(personal_time),
            "Mean Extracurricular Time" = mean(extracurricular_time),
            "Median Extracurricular Time" = median(extracurricular_time),
            "Standard Deviation (Extracurricular time)" = sd(extracurricular_time)) %>%
#Flipping the columns to rows and rows to columns to make the table nicer
  t() %>%
#Shifting the first row to be the column names
  janitor::row_to_names(1) %>%
#t() turned the data into a matrix, putting it back in a data frame format
  as.data.frame() %>%
  mutate(Weekday = round(as.numeric(Weekday), 2), 
         Weekend = round(as.numeric(Weekend), 2))

  






```



# Results 

<!--
You may organize your results however you see fit (e.g., one subsection for each question or one subsection for each visualization/table). 

Each visualization and table should have its own code chunk. 

Before each code chunk, describe what information is conveyed through the corresponding visualization or table in connection with the questions you posed (e.g., "The first figure captures how the time I spent on each class varied over the two-week period. It appears I spent the most time on..."). 
-->


<!--
Describe what information is conveyed through your first data visualization (that you'll create below).

-->
My first visualization shows the average amount of time I spend on each class per day of the week. The x axis represents a particular day of the week (ie. monday, sunday) and the y axis represents the average amount of time I spend in hours on an activity. A line is used for each class, so that each point on the line at a particular day represents the average amount of time I spend on that class on that day. 

This visualization reveals that I spend the most time on data science towards the end of the week, and the most time on religion towards the beginning of the week. Overall, I seem to spend less time on oceanography than my other classes and the bulk of that work is still concentratedtowards the beginning of the week. Finally, time spent on my thesis is scattered throughout the week with no clear peak.

```{r visualization-1, eval = TRUE}
# Code for first data visualization
p <- ggplot(data = activities_average, 
            aes(x=day_of_week, 
                y = Duration,
                color = Activity, 
                group = Activity)) +
#Setting the size for the lines
  geom_line(size = 1.2, alpha = 1) + 
#Setting line colors and labels for the legend
  scale_color_brewer(type = "qual", 
                     palette = 7, 
                     labels = c("Data Science", 
                                "Oceanography", 
                                  "Apocalyptic Religion", 
                                  "Thesis")) +
#Setting graph title and axes labels
                     labs(DS_Average = "Data Science", 
                          title = "Average Work Time Per Class Over a Week",
                          x = "Day", 
                          y = "Average Work Time (hours)",
                          color = "Class") +
#Adjusting font size
  theme(legend.text = element_text(size = 8)) +
  theme(legend.title = element_text(size = 10))

p

```

<!--
Describe what information is conveyed through your second data visualization (that you'll create below).

-->

My second visualization shows the relationship between the amount of time I work per day to the amount of time I spend on personal time per day. The x axis represents the amount of work per day, and I was interested in how this impacted the amount of personal time I took, which I put on the y axis. I used faceting to divide this data into two separate graphs for the weekend and weekday. 

It appears that there is no clear relationship between the amount of work time I take in a day and the amount of personal time I take in a day, lending support to neither of my hypotheses. Because of this lack of a clear relationship, I did not add a line of best fit to the graphs as it would not be appropriate. It does appear that there may be a slightly negative relationship between personal time and work time on the weekends, but I would need more data to draw any significant conclusion.


```{r visualization-2, eval = TRUE}
#Plotting total work time against personal time per day
n <- ggplot(data = activities_comparison2, aes(x= work_time, y = personal_time)) +
#Creating 1 plot for weekends and one for weekdays
  facet_wrap(~week_status, nrow = 1) +
#Using points to represent each day
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
#Labeling the graph
  labs(title = "Time spent on work vs Recreation per day",
       y = "Personal time (hours)",
       x = "Schoolwork time (hours)")
n
```

<!--
Describe what information is conveyed through the table (that you'll create) below. Again, the order is up to you. Move these around so they make sense for your analysis!
-->

This table provides summary statistics about the ways I spend my time on the weekdays and the weekend. Specifically, it indicates the mean, median, and standard deviation of my schoolwork time, personal time, and extracurricular time spent on weekdays vs. weekends. The table reveals that I spend more time on schoolwork during weekdays, and take more personal time on weekends. No noticeably different patterns were present for extracurricular time.


```{r, results = 'asis', eval = TRUE}
# Code for table
# Only code for your table should be here (no or very minimal wrangling code)
week_comparison %>%
#Titling the table, setting digit cap
  kable(booktabs = TRUE, digits = 3, caption = "Differences in Time Spent on Weekdays vs. Weekends") %>%
#Holding the table in place on the pdf
  kable_styling(latex_options = "HOLD_position")




```

# Conclusions 

<!--
With your initial questions in mind, briefly summarize what you learned about how you spent your time. What are the big take-aways for you about how you spend your time? If relevant, how does this affect how you might spend your time going forward?
-->

I learned that I have succeeded fairly well in my goal of planning out my week such that I do work for each class on a pre-planned part of the week. I tend to do most of my data science work at the end of the week, my religion & oceanography work at the start of the week, and my thesis work scattered throughout the week and weekend. I also learned in this analysis that I spend the most time on data science rather than my thesis, which was somewhat surprising. However, I did collect my data during two non-representative weeks, something I will address in my reflection. 

Next, I learned that there is no strong relationship between how much time I spend on work and how much personal time I take in a day. The idea that I give myself a break when I do a lot of work was not confirmed by the data, suggesting that I should not count on this pattern when planning out my days.

Finally, I learned that I do seem to spend more time on work during the weekdays and give myself a greater break on the weekends. During the weekdays I work an average of approximately 5 hours and this reduces to approximately 3 on the weekends. The average amount of personal time I take on the weekends is approximately double that of the weekdays. 




\newpage


# Reflection 

<!--
Write your reflection here in paragraph form. In particular, address...

* What difficulties in the data collection and analysis process did you encounter?  Identify two of your main hurdles in gathering accurate data.
* What implications do those difficulties have for future data collection and/or analysis projects if you were to repeat this project? 
* Did you have enough data? How much data do you think you'd need to collect in order to satisfactorily answer your question(s) of interest? Would it be hard to collect that data? Why or why not?
* As someone who provides data, what expectations do you have when you give your data (e.g. to Facebook, Google, MapMyRun, etc.)?  
* As someone who analyzes others' data, what ethical responsibilities do you have?

You may address these questions in any particular order. If it helps, you can make a subsection for each. 
-->

Difficulties & Future data projects




One difficulty I encountered when analyzing my data is that I did not code all activities I did, only a certain set of activities. This was a problem because my second analysis involved looking at how the amount of schoolwork I did in a day affected the personal time I took, and this would have been neater if I was able to compare it to the amount of non-college related work I did. If I were to repeat this project, I would have coded in not only personal time but the time I spent on non-school work like independent projects or looking for jobs. Without having explicitly coded in my non-school work time, I was only able to indirectly measure that time via personal time, which was not ideal.

Another difficulty I encountered was in the process of coding data. Initially I was not consistently calling each activity the same label. For instance, I might use 'oceanography class', 'homework', and 'oceanography studying' as labels on different days. If I wanted to dig into each of these specific categories this would have been fine, but I wanted to analyze all these events as one category and R recognizes them separately. I could have either used wrangling to do this or simply recoded my information, and I ended up choosing the latter. If I could repeat the project I would be more careful to come up with a distinct system for coding ahead of time where I know exactly what categories I want to examine, and then been sure to name each of these timeslots the exact same thing.



Did I have enough data?

I think I did not have quite enough data to completely answer some of my important questions. For instance, my first question was about how the time I spend on each class changed throughout the course of an average week. However, I only had two weeks worth of data to average across. These two weeks are not perfectly representative of my average week. For instance, I have spent a lot more time on my thesis in the past and happened to have less work for it in the last two weeks since I am between writing sections. To get a representative sample of my typical week I think collecting data for at least 1-2 months would be ideal. Collecting this data would not be too difficult since I would just need to continue to track my time for a full month. I think the same point applies to my analysis of work time on weekdays vs. weekends, it would be better to have a more representative sample by collecting data for a few months.



Providing Data

I expect that when I provide data that my personal information will be anonymized. It would be a violation of my privacy for companies to take sensitive personal information and share it with others without my consent. I also expect that my data will be used for ethically acceptable purposes. For instance, it would be concerning if companies sold my data to more malicious corporations that wanted to target particular people with false advertising they may be susceptible to. Thus, I expect that there is some system by which we grant explicit consent for our data to be used and background principles that this data only be shared for useful & ethically acceptable purposes.

Ethical Responsibilities

When analyzing others data, I have the responsibility to check whether I have permission to use their data and to respect the privacy of the people referenced by the data whenever possible. For instance, we should anonymize data for people who are uncomfortable with their personal information being shared in data analyses. I also have the responsibility to work to mitigate biased analyses wherever possible. This means not only avoiding analyses that use factors such as race or gender in unscrupulous ways, but also being vigilant in avoiding 'proxies' for these categories that end up producing results based on the legacy of discrimination in this country.

